\large{\bf Looking forwards and backwards: dynamics and genealogies of locally regulated populations }

### 2023-12-11 forwards-and-backwards.tex
.

@@

Reproduction of individuals naturally leads to spatial branching process models,
including branching random walk, 
branching Brownian motion, and the Dawson-Watanabe superprocesses.  
However, as a result of the branching assumption (once born, individuals behave 
independently of one another), 
a population evolving according to any of these models will
either die out or grow without bound and, in so doing, can
develop clumps of arbitrarily large density and extent. 
Our starting point here is an individual-based model of a single species 
in continuous space in which
birth, death, and establishment may all depend on local population density
as well as on spatial location,
allowing for stable populations through density-dependent feedback.

@@

To understand the stability of the constant solution $\varphi\equiv\varphi_0$, 
we
linearise~\eqref{general deterministic limit} around $\varphi_0$:
let $\varphi_t(x) = \varphi_0 + \psi_t(x)$,
and (informally) $r(x) \approx r(\varphi_0) + r'(\varphi_0) \smooth{r} \psi(x)$.
Writing $r_0 = r(\varphi_0)$ and $r'_0 = r'(\varphi_0)$,
with analogous expressions for $\gamma$ and $F$,
\begin{align*}
    \partial_t \psi
    &\approx
    \sigma^2 \varphi_0 r_0 \gamma_0' \Delta \smooth{\gamma} \psi
    + \sigma^2 r_0 \gamma_0 \Delta \psi
    + \varphi_0 F'_0 \smooth{F} \psi .
\end{align*}
Letting $\hat f(u) = \int e^{i u x} f(x) dx / \sqrt{2 \pi}$ denote the Fourier transform,
\begin{equation}
	\label{fourier transform psi}
    \partial_t \hat \psi(u)
    \approx
	\left\{
	- u^2\sigma^2 \varphi_0 r_0 \gamma_0' \hat{\rho}_\gamma(u) 
	%e^{-\epsilon^2 u^2 /2}
        - u^2 \sigma^2 r_0 \gamma_0
	+ \varphi_0 F'_0 \hat{\rho}_F(u) 
	%e^{-\epsilon^2 u^2 /2} 
	\right\} \hat \psi(u) .
\end{equation}
In the simplest case, in which $\gamma$ is constant, so $\gamma_0'=0$,
this reduces to
\begin{equation} %\begin{align*}
	\label{simplest FT}
	\partial_t \hat \psi(u)
    \approx
    \left(
%	- u^2\sigma^2 \varphi_0 r_0 \gamma_0' \hat{\rho}_\gamma(u) 
	%e^{-\epsilon^2 u^2 /2}
        - u^2 \sigma^2 r_0 \gamma_0
	+ \varphi_0 F'_0 \hat{\rho}_F(u) 
	%e^{-\epsilon^2 u^2 /2} 
    \right) \hat \psi(u) .
\end{equation} %\end{align*}
If we take $\rho_F=p_{\epsilon^2}$, then 
$\hat{\rho}_F(u)=\exp(-\epsilon^2 u^2 / 2) / \sqrt{2 \pi}$ and (recalling 
that $F_0'<0$) the term in brackets is always negative, and we recover
the well-known fact that in this case the constant solution is stable.
If, on the other hand, $\hat{\rho}_F$ changes sign, there may be values of $u$ for which the corresponding quantity is positive. For example, if $d=1$ and
$\rho_F(x)={\mathbf 1}_{[-\epsilon, \epsilon]}(x)/2\epsilon$, then 
$\hat{\rho}_F(u)=\sin(\epsilon u)/(\sqrt{2\pi}\epsilon u)$, which is 
negative for $u\in (\pi/epsilon, 2\pi/\epsilon)$ (and periodically 
repeating intervals). Setting $v=\epsilon u$, the bracketed term on 
the right hand side of~\eqref{simplest FT} becomes 
\[
    \varphi_0F_0'\frac{1}{\sqrt{2\pi} v} \sin(v)
	-\frac{\sigma^2}{\epsilon^2}v^2r_0\gamma_0,
\]
and we see that if $\sigma^2/\epsilon^2$ is sufficiently small, there are
values of $v$ for which this is positive. In other words, in keeping
with our heuristic above, if dispersal is sufficiently short range relative to
the range over which individuals interact, there are unstable frequencies that
scale with the interaction distance $\epsilon$. In two dimensions, replacing 
the indicator of an interval by that of a ball of radius $\epsilon$, a
similar analysis applies, except that the sine function is replaced by a 
Bessel function.

@@

\begin{lemma}
\label{no mass at infty}
Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
if $(\eta_t)_{t\geq 0}$ be a limit point of $\{(\eta^N_t)_{t\geq 0}\}_{N\geq 1}$,
then for any $\delta>0$ and $T>0$,
\[
\IP\bigg[
    \sup_{0 \le t \le T} \eta_t\big(\{\|x\|>R\}\big)>\delta
\bigg]\to 0\qquad\mbox{ as }R\to\infty.
\]
\end{lemma}

>>>>

\comment{TODO: double-check $\langle \cdot \rangle$ versus $[\cdot]$}
\comment{TODO: check factor of 2}

@@

Reproduction of individuals naturally leads to spatial branching process models,
including branching random walk, 
branching Brownian motion, and the Dawson-Watanabe superprocesses.  
However, as a result of the branching assumption (once born, individuals behave 
independently of one another), 
a population evolving according to any of these models will
either die out or grow without bound and, in so doing, can
develop clumps of arbitrarily large density and extent. 
Our starting point here is an individual-based model of a single species 
in continuous space in which
birth, death, and establishment may all depend on local population density
as well as on spatial location,
allowing for stable populations through density-dependent feedback.
The model generalizes those introduced to the ecology literature by
\citet{bolker/pacala:1997} and \citet{law/murrell/dieckmann:2003},
and subsequently studied, e.g., by \citet{etheridge:2004} and \citet{fournier2004microscopic}.

@@

To understand the stability of the constant solution $\varphi\equiv\varphi_0$, 
we linearise~\eqref{general deterministic limit} around $\varphi_0$:
let $\varphi_t(x) = \varphi_0 + \psi_t(x)$,
and (informally) $r(x) \approx r(\varphi_0) + r'(\varphi_0) \smooth{r} \psi(x)$.
Recall that in this section we are in $d=1$.
Writing $r_0 = r(\varphi_0)$ and $r'_0 = r'(\varphi_0)$,
with analogous expressions for $\gamma$ and $F$,
\begin{align*}
    \partial_t \psi
    &\approx
    \sigma^2 \varphi_0 r_0 \gamma_0' \Delta \smooth{\gamma} \psi
    + \sigma^2 r_0 \gamma_0 \Delta \psi
    + \varphi_0 F'_0 \smooth{F} \psi .
\end{align*}
Letting $\hat f(u) = \int e^{i u x} f(x) dx / \sqrt{2 \pi}$ denote the Fourier transform,
\begin{equation}
	\label{fourier transform psi}
    \partial_t \hat \psi(u)
    \approx
	\left\{
	- u^2\sigma^2 \varphi_0 r_0 \gamma_0' \hat{\rho}_\gamma(u) 
	%e^{-\epsilon^2 u^2 /2}
        - u^2 \sigma^2 r_0 \gamma_0
	+ \varphi_0 F'_0 \hat{\rho}_F(u) 
	%e^{-\epsilon^2 u^2 /2} 
	\right\} \hat \psi(u) .
\end{equation}
In the simplest case, in which $\gamma$ is constant, so $\gamma_0'=0$,
this reduces to
\begin{equation} %\begin{align*}
	\label{simplest FT}
	\partial_t \hat \psi(u)
    \approx
    \left(
%	- u^2\sigma^2 \varphi_0 r_0 \gamma_0' \hat{\rho}_\gamma(u) 
	%e^{-\epsilon^2 u^2 /2}
        - u^2 \sigma^2 r_0 \gamma_0
	+ \varphi_0 F'_0 \hat{\rho}_F(u) 
	%e^{-\epsilon^2 u^2 /2} 
    \right) \hat \psi(u) .
\end{equation} %\end{align*}
If we take $\rho_F=p_{\epsilon^2}$, then 
$\hat{\rho}_F(u)=\exp(-\epsilon^2 u^2 / 2) / \sqrt{2 \pi}$ and (recalling 
that $F_0'<0$) the term in brackets is always negative, and we recover
the well-known fact that in this case the constant solution is stable.
If, on the other hand, $\hat{\rho}_F$ changes sign, there may be values of $u$ for which the corresponding quantity is positive. For example, if $d=1$ and
$\rho_F(x)={\mathbf 1}_{[-\epsilon, \epsilon]}(x)/2\epsilon$, then 
$\hat{\rho}_F(u)=\sin(\epsilon u)/(\sqrt{2\pi}\epsilon u)$, which is 
negative for $u\in (\pi/\epsilon, 2\pi/\epsilon)$ (and periodically 
repeating intervals). Setting $v=\epsilon u$, the bracketed term on 
the right hand side of~\eqref{simplest FT} becomes 
\[
    \varphi_0F_0'\frac{1}{\sqrt{2\pi} v} \sin(v)
	-\frac{\sigma^2}{\epsilon^2}v^2r_0\gamma_0,
\]
and we see that if $\sigma^2/\epsilon^2$ is sufficiently small, there are
values of $v$ for which this is positive. In other words, in keeping
with our heuristic above, if dispersal is sufficiently short range relative to
the range over which individuals interact, there are unstable frequencies that
scale with the interaction distance $\epsilon$. In two dimensions, replacing 
the indicator of an interval by that of a ball of radius $\epsilon$, a
similar analysis applies, except that the sine function is replaced by a 
Bessel function.

@@

\begin{lemma}
\label{no mass at infty}
Under the assumptions of Theorem~\ref{thm:nonlocal_convergence},
if $(\eta_t)_{t\geq 0}$ is a limit point of $\{(\eta^N_t)_{t\geq 0}\}_{N\geq 1}$,
then for any $\delta>0$ and $T>0$,
\[
\IP\bigg[
    \sup_{0 \le t \le T} \eta_t\big(\{\|x\|>R\}\big)>\delta
\bigg]\to 0\qquad\mbox{ as }R\to\infty.
\]
\end{lemma}
